TacticToe: Learning the Game of Tactics

### 2018-02-01 tactictoe_jv.tex
recording: wip

\begin{itemize}
\item Proof recording at the tactic level is made more precise. Pattern 
matching constructions, opening of submodules are supported. The different 
recording steps (parsing, unfolding and replaying) are regrouped into a single 
\sml function.
\item Monte Carlo tree search~~\cite{montecarlo} (MCTS) replaces A* as our 
search algorithm. The MCTS algorithm gives a more balanced feedback on 
the success of 
each proof step by
comparing subtrees of the search tree instead of leafs. The policy and 
evaluation are learned 
through supervised learning.
\item Proof guidance required by MCTS is given by predictors for 
three kind of objects: tactics, theorems and list of goals.
\item The orthogonalization process that eliminates redundant tactics is 
formally described.
\item A tactic abstraction method is introduced. It enable us to create 
more general and flexible tactics where tactic arguments are dynamically 
predicted.
\item The internal ATP \metis is complemented 
by asynchronous calls to \eprover to help \tactictoe during proof search.
\todoi{More on evaluation??}
\item Evaluation of \tactictoe success rate relative to 
the length of the original proofs.
\item User interaction is enhanced by the minimization and prettification 
applied on the generated proof.
\end{itemize}

@@

\paragraph{Datasets: optimization and validation}\todo{rephrase to something 
more 
coherent}
All top-level theorems from the standard library are considered with the 
exception of 440 hard problems (containing a \texttt{let} construction in their 
proof) and 1175 easy problems (build from \texttt{save\_thm} calls). 
Therefore, during the full-scale experiments, we evaluate 7164 theorems.
We use every tenth theorem of the first third of the standard library of 
them for parameter optimization, which amounts to 273 theorems.
Although the evaluation of each set of parameters on its own is fair, 
the selection of the best strategy in Section~\ref{sec:tuning} should also be 
considered as a learning process. To ensure the global fairness, the final 
experiments in Section~\ref{sec:full_exp} runs the best strategy on the full 
dataset which is about 30 times larger.

@@

Rather than relying on the underlying programming language, we parse the
original theory file containing proof scripts. This enables us to extract the 
code of each tactic. 
Working with the string representation of a tactic is better that working with 
its value:
\begin{itemize}
\item Equality between two functions is easy to do from 
their string representation. The direct consequence is that we can avoid 
predicting the same tactic over and over again. 
\item 
The 
results of 
\tactictoe can be returned in a readable form to the user. In this way, the 
user can learn from the feedback, analyze the proof and possibly improve on it.
An other point is that \tactictoe does not need to be installed to run  
proof scripts generated by \tactictoe. In this way, the further development of 
\tactictoe are not affecting the robustness of \holfour.
\item
It is probably difficult (even impossible) to transfer values 
between theories as they are not run in the same \holfour session. In contrast, 
\sml code can be exported, imported.
\item To produce a tactic value form a code is easy in \sml, which can be 
achieved by using a reference to a tactic and update it with the function 
\texttt{use}.
\end{itemize}

@@

In the next phase, we globalize identifiers of the proof scripts. 
Infix operators such as \texttt{THEN} need to be processed in a special way so 
that they can keep their infixity after globalization. For simplicity, the 
globalization of infix operators is omitted in Example~\ref{ex:r1}.
In this example, the three main cases that can happen during the globalization
are depicted. The first one is the globalization of identifiers declared in 
modules. The global identifiers \texttt{STRIP\_TAC} and 
\texttt{ASM\_REWRITE\_TAC} are prefixed by their module \texttt{Tactic}. In 
this way, they will still be interpretable whether \texttt{Tactic} was open or 
not. The local identifier \texttt{LIST\_INDUCT\_TAC} is replaced by its 
definition which happens to contain two global identifiers.
The previous paragraph describe the globalization for all identifiers except 
local theorems.  We do not replace a local
theorem by its definition because we want to avoid unfolding proof scripts into 
other proof scripts.
If the theorem is stored in the database of the current theory, we 
can obtain the theorem value by calling \texttt{DB.fetch}. This happens for the 
three theorems in our script. The internal database of \holfour is available 
across \holfour developments. 
Otherwise the globalization process fails and the local theorem identifier is 
kept unchanged. A recorded tactic with an unchanged local theorem as argument 
is only interpretable inside the current theory.

@@

\todo{rewrite this paragraph}

The intended effect of the splitting function is to detect and wrap \sml 
expressions of type tactics. Its aim is to is encapsulate each tactic 
unit $t$ within a recording function $R$. This function $R$ is designed to 
observe what goals $t$ receives at run time. Since there can be maybe nested 
level of tactics, (i.e 
tactics that contain tactics as subexpressions), we choose one that is
close to what a human understand to be a tactic unit. 
Most tacticals are infix, and we do not want to consider combination of 
tactics to be a tactic unit. For \texttt{Split}, a tactic unit is a 
subterm of 
type tactics that do not contain a infix operator as its top symbol. Thanks to
globalized infixity, we just have to verify that the expression 
is of type tactic and that an infix guard is not a top symbol of the 
expression.
To determine that, we run the \polyml compiler on \texttt{"p'"} to give us a 
detailed code tree of the expression. Each tactic unit is then substituted 
by a wrapped version to produce the final proof script.

@@

\begin{example}\label{ex:wrap} (Tactic unit wrapping)
\begin{lstlisting}[language=SMLSmall,frame=tb]
R "Tactic.STRIP_TAC" THEN
R "Prim_rec.INDUCT_THEN (DB.fetch \"list\" \"INDUCT\") Tactic.ASSUME_TAC" THEN
R "Rewrite.ASM_REWRITE_TAC 
  [fff \"list\" \"MAP\", fff \"list\" \"APPEND\"]"
\end{lstlisting}
\end{example}

\todo{rewrite this paragraph}

During its execution, the recording function \texttt{R} (as coded in 
Definition~\ref{ex:record}) 
saves the tactic 
string and the goal, verifies that the tactic string as the same effect as the 
original tactic and apply it to continue the proof. To perform this task, we 
convert the tactic string back into a tactic via the function 
\texttt{tactic\_of\_sml}.

\begin{example}\label{ex:record} (Code of the recording function)
\begin{lstlisting}[language=SMLSmall,frame=tb]
fun R stac goal = (save (stac,goal); tactic_of_sml stac goal)
\end{lstlisting}
\end{example}

The function \texttt{save} writes the tactic-goal pairs to disk increasing the 
number of entries in our database of tactics. The function 
\texttt{tactic\_of\_sml} interprets \sml code \texttt{stac}. The tactic is then 
applied to the input goal which replicates the original behavior.

@@

Goal: \texttt{IS\_GCD\_UNIQUE} in theory \texttt{pred\_set}).
\begin{lstlisting}[language=SMLSmall,frame=tb]
!a b c d. is_gcd a b c wedge is_gcd a b d ==> (c = d)
\end{lstlisting}

Original \holfour proof script
\begin{lstlisting}[language=SMLSmall,frame=tb]
PROVE_TAC[IS_GCD, DIVIDES_ANTISYM]
\end{lstlisting}

@@

\tactictoe proof script
\begin{lstlisting}[language=SMLSmall,frame=tb]
STRIP_TAC THEN 
REWRITE_TAC [fetch "gcd" "is_gcd_def"] THEN 
REPEAT Cases THENL 
  [METIS_TAC [], 
   REWRITE_TAC [SUC_NOT, ALL_DIVIDES_0, compute_divides] THEN 
     METIS_TAC [NOT_SUC], 
   METIS_TAC [NOT_SUC, DIVIDES_ANTISYM], 
   METIS_TAC [LESS_EQUAL_ANTISYM, DIVIDES_LE, LESS_0], 
   METIS_TAC [], 
   RW_TAC numLib.arith_ss [divides_def], 
   METIS_TAC [DIVIDES_ANTISYM],
   METIS_TAC [LESS_EQUAL_ANTISYM, DIVIDES_LE, LESS_0]]
\end{lstlisting}

@@

Goal: \texttt{SURJ\_INJ\_INV} in theory \texttt{pred\_set})
\begin{lstlisting}[language=SMLSmall,frame=tb]
!f s t. SURJ f s t ==> ?g. INJ g t s wedge !y. y IN t ==> (f (g y) = y)
\end{lstlisting}

Original \holfour proof script (2 milliseconds)
\begin{lstlisting}[language=SMLSmall,frame=tb]
REWRITE_TAC [IMAGE_SURJ] THEN
DISCH_TAC THEN Q.EXISTS_TAC `THE o LINV_OPT f s` THEN
BasicProvers.VAR_EQ_TAC THEN REPEAT STRIP_TAC THENL 
  [irule INJ_COMPOSE THEN Q.EXISTS_TAC `IMAGE SOME s` THEN
     REWRITE_TAC [INJ_LINV_OPT_IMAGE] THEN REWRITE_TAC [INJ_DEF, IN_IMAGE] THEN
     REPEAT STRIP_TAC THEN REPEAT BasicProvers.VAR_EQ_TAC THEN
     FULL_SIMP_TAC std_ss [THE_DEF],
   ASM_REWRITE_TAC [LINV_OPT_def, o_THM, THE_DEF] THEN
     RULE_ASSUM_TAC (Ho_Rewrite.REWRITE_RULE
       [IN_IMAGE', GSYM SELECT_THM, BETA_THM]) THEN ASM_REWRITE_TAC []]
\end{lstlisting}

@@

Make it match the new version.
\tactictoe proof script (50 milliseconds)
\begin{lstlisting}[language=SMLSmall,frame=tb]
SRW_TAC [] [SURJ_DEF, INJ_DEF] THEN METIS_TAC []
\end{lstlisting}

Goal
\begin{lstlisting}[language=SMLSmall,frame=tb]
!l1 l2 n x. LENGTH l1 <= n ==> 
  (LUPDATE x n (l1 ++ l2) = l1 ++ (LUPDATE x (n - LENGTH l1) l2))
\end{lstlisting}

@@

Original \holfour proof script (63 milliseconds)
\begin{lstlisting}[language=SMLSmall,frame=tb]
  rw[] THEN simp[LIST_EQ_REWRITE] THEN Q.X_GEN_TAC `z` THEN
  simp[EL_LUPDATE] THEN rw[] THEN simp[EL_APPEND2,EL_LUPDATE] THEN
  fs[] THEN Cases_on `z < LENGTH l1` THEN
  fs[] THEN simp[EL_APPEND1,EL_APPEND2,EL_LUPDATE]
\end{lstlisting}

@@

%val LUPDATE_APPEND2 = Q.store_thm("LUPDATE_APPEND2",
%   `!l1 l2 n x.
%      LENGTH l1 <= n ==>
%      (LUPDATE x n (l1 ++ l2) = l1 ++ (LUPDATE x (n-LENGTH l1) l2))`,
%);
\vspace{5mm}
Make it match the new version.
\tactictoe proof script (17 milliseconds)
\begin{lstlisting}[language=SMLSmall,frame=tb]
Induct_on `l1` THENL [SRW_TAC [] [],
  Cases_on `n` THENL [SRW_TAC [] [],
    FULL_SIMP_TAC (srw_ss ()) [] THEN METIS_TAC [LUPDATE_def]]]
\end{lstlisting}
fs is a \sml binding for FULL\_SIMP\_TAC (srw\_ss ()) and rw is short for ??.

>>>>

\begin{itemize}
\item Proof recording at the tactic level is made more precise. Pattern 
matching constructions, opening of submodules are supported. The different 
recording steps (parsing, unfolding and replaying) are regrouped into a single 
\sml function.
\item Monte Carlo tree search~~\cite{montecarlo} (MCTS) replaces A* as our 
search algorithm. The MCTS algorithm gives a more balanced feedback on 
the success of 
each proof step by
comparing subtrees of the search tree instead of leafs. The policy and 
evaluation are learned 
through supervised learning.
\item Proof guidance required by MCTS is given by predictors for 
three kind of objects: tactics, theorems and list of goals.
\item The orthogonalization process that eliminates redundant tactics is 
formally described.
\item A tactic abstraction method is introduced. It enables us to create 
more general and flexible tactics where tactic arguments are dynamically 
predicted.
\item The internal ATP \metis is complemented 
by asynchronous calls to \eprover to help \tactictoe during proof search.
\todoi{More on evaluation??}
\item Evaluation of \tactictoe success rate relative to 
the length of the original proofs.
\item User interaction is enhanced by the minimization and prettification 
applied on the generated proof.
\end{itemize}

@@

\todoi{rephrase to something more coherent}
\paragraph{Datasets: optimization and validation}

All top-level theorems from the standard library are considered with the 
exception of 440 hard problems (containing a \texttt{let} construction in their 
proof) and 1175 easy problems (build from \texttt{save\_thm} calls). 
Therefore, during the full-scale experiments, we evaluate 7164 theorems.
We use every tenth theorem of the first third of the standard library of 
them for parameter optimization, which amounts to 273 theorems.
Although the evaluation of each set of parameters on its own is fair, 
the selection of the best strategy in Section~\ref{sec:tuning} should also be 
considered as a learning process. To ensure the global fairness, the final 
experiments in Section~\ref{sec:full_exp} runs the best strategy on the full 
dataset which is about 30 times larger.

@@

Rather than relying on the underlying programming language, we parse the
original theory file containing proof scripts. This enables us to extract the 
code of each tactic. 
Working with the string representation of a tactic is better that working with 
its value:
\begin{itemize}
\item Equality between two functions is easy to do from 
their string representation. The direct consequence is that we can avoid 
predicting the same tactic over and over again. 
\item 
The 
results of 
\tactictoe can be returned in a readable form to the user. In this way, the 
user can learn from the feedback, analyze the proof and possibly improve on it.
An other point is that \tactictoe does not need to be installed to run  
proof scripts generated by \tactictoe. In this way, the further development of 
\tactictoe are not affecting the robustness of \holfour.
\item
It is difficult (probably impossible) to transfer \sml values 
between theories as they are not run in the same \holfour session. In contrast, 
\sml code can be exported, imported.
\item To produce a tactic value form a code is easy in \sml, which can be 
achieved by using a reference to a tactic and update it with the function 
\texttt{use}.
\end{itemize}

@@

In the next phase, we globalize identifiers of the proof scripts. 
Infix operators such as \texttt{THEN} need to be processed in a special way so 
that they keep their infixity status after globalization. For simplicity, 
the 
globalization of infix operators is omitted in Example~\ref{ex:r1}.
In this example, the three main cases that can happen during the globalization
are depicted. The first one is the globalization of identifiers declared in 
modules. The global identifiers \texttt{STRIP\_TAC} and 
\texttt{ASM\_REWRITE\_TAC} are prefixed by their module \texttt{Tactic}. In 
this way, they will still be interpretable whether \texttt{Tactic} was open or 
not. The local identifier \texttt{LIST\_INDUCT\_TAC} is replaced by its 
definition which happens to contain two global identifiers.
The previous paragraph describe the globalization for all identifiers except 
local theorems.  We do not replace a local
theorem by its definition because we want to avoid unfolding proof scripts into 
other proof scripts.
If the theorem is stored in the database of the current theory, we 
can obtain the theorem value by calling \texttt{DB.fetch}. This happens for the 
three theorems in our script. The internal database of \holfour theorems is 
available across \holfour developments. 
Otherwise the globalization process fails and the local theorem identifier is 
kept unchanged. A recorded tactic with an unchanged local theorem as argument 
is only interpretable inside the current theory.

@@

Running the globalized version of a proof script will have the exact same 
effect as the original. But since we want to extract information from this 
script in the form tactics and their input goals, we modify it further.
In particular, we need to define at which level we should record the tactics of 
the proof scripts. The simplest idea would be to record all \sml subexpressions 
of type tactics. However, it will damage the quality of our data by 
dramatically increasing the number of tactic associated with a single goal.
Imagine a proof script of the form \texttt{A THEN B THEN C} then the tactics
\texttt{A}, \texttt{A THEN B} and \texttt{A THEN B THEN C} would be valid 
advice for something close to their common input goal. The tactic  
\texttt{A THEN B THEN C} is likely to be specific. In contrast, we can consider 
the tactic \texttt{REPEAT A} that is repetitively calling the tactic \texttt{A} 
until 
\texttt{A} has no effect. In this situation, it is often preferable to record 
only \texttt{REPEAT A} and any call to \texttt{A}. Otherwise during the 
proof search, we would have to make a branches created after each call of 
\texttt{A}. 
To sum up, we would like to record only the most general tactics which 
make the most progress on a goal. As a trade-off between these two objectives,
we split the proofs into tactic units.

@@

Because these tactic units are constructed from visual information present in 
the proof script, they often represent what a human consider to be one step of 
the proof.

To produce the final recording script, we encapsulate each tactic unit 
in a recording function \texttt{R} (see Example~\ref{ex:wrap}). In order to 
record 
tacitcs in all \holfour theories, we replace the original proof by the 
recording script in each theory and rebuild the \holfour library.

\begin{example}\label{ex:wrap} (Tactic unit wrapping)
\begin{lstlisting}[language=SMLSmall]
R "Tactic.STRIP_TAC" THEN
R "Prim_rec.INDUCT_THEN (DB.fetch \"list\" \"INDUCT\") Tactic.ASSUME_TAC" THEN
R "Rewrite.ASM_REWRITE_TAC 
  [fff \"list\" \"MAP\", fff \"list\" \"APPEND\"]"
\end{lstlisting}
\end{example}

At run time the function \texttt{R} is designed to observe what input goals a
tactic  receives without changing its output. The implementation of \texttt{R} 
is presented in Example~\ref{ex:record}.

\begin{example}\label{ex:record} (Code of the recording function)
\begin{lstlisting}[language=SMLSmall]
fun R stac goal = (save (stac,goal); tactic_of_sml stac goal)
\end{lstlisting}
\end{example}

The function \texttt{save} writes the tactic-goal pair to disk increasing the 
number of entries in our database of tactics. The function 
\texttt{tactic\_of\_sml} interprets the \sml code \texttt{stac}. The tactic is 
then applied to the input goal to replicate the original behavior.
After all modified theories are rebuild, each call to a wrapped tactic in a
proof script is recorded as a pair containing its globalized code and its 
input goal.

@@

Goal: \texttt{IS\_GCD\_UNIQUE} in theory \texttt{pred\_set}).
\begin{lstlisting}[language=SMLSmall]
!a b c d. is_gcd a b c wedge is_gcd a b d ==> (c = d)
\end{lstlisting}

Original \holfour proof script
\begin{lstlisting}[language=SMLSmall]
PROVE_TAC[IS_GCD, DIVIDES_ANTISYM]
\end{lstlisting}

@@

\tactictoe proof script
\begin{lstlisting}[language=SMLSmall]
STRIP_TAC THEN 
REWRITE_TAC [fetch "gcd" "is_gcd_def"] THEN 
REPEAT Cases THENL 
  [METIS_TAC [], 
   REWRITE_TAC [SUC_NOT, ALL_DIVIDES_0, compute_divides] THEN 
     METIS_TAC [NOT_SUC], 
   METIS_TAC [NOT_SUC, DIVIDES_ANTISYM], 
   METIS_TAC [LESS_EQUAL_ANTISYM, DIVIDES_LE, LESS_0], 
   METIS_TAC [], 
   RW_TAC numLib.arith_ss [divides_def], 
   METIS_TAC [DIVIDES_ANTISYM],
   METIS_TAC [LESS_EQUAL_ANTISYM, DIVIDES_LE, LESS_0]]
\end{lstlisting}

@@

Goal: \texttt{SURJ\_INJ\_INV} in theory \texttt{pred\_set})
\begin{lstlisting}[language=SMLSmall]
!f s t. SURJ f s t ==> ?g. INJ g t s wedge !y. y IN t ==> (f (g y) = y)
\end{lstlisting}

Original \holfour proof script (2 milliseconds)
\begin{lstlisting}[language=SMLSmall]
REWRITE_TAC [IMAGE_SURJ] THEN
DISCH_TAC THEN Q.EXISTS_TAC `THE o LINV_OPT f s` THEN
BasicProvers.VAR_EQ_TAC THEN REPEAT STRIP_TAC THENL 
  [irule INJ_COMPOSE THEN Q.EXISTS_TAC `IMAGE SOME s` THEN
     REWRITE_TAC [INJ_LINV_OPT_IMAGE] THEN REWRITE_TAC [INJ_DEF, IN_IMAGE] THEN
     REPEAT STRIP_TAC THEN REPEAT BasicProvers.VAR_EQ_TAC THEN
     FULL_SIMP_TAC std_ss [THE_DEF],
   ASM_REWRITE_TAC [LINV_OPT_def, o_THM, THE_DEF] THEN
     RULE_ASSUM_TAC (Ho_Rewrite.REWRITE_RULE
       [IN_IMAGE', GSYM SELECT_THM, BETA_THM]) THEN ASM_REWRITE_TAC []]
\end{lstlisting}

@@

Make it match the new version.
\tactictoe proof script (50 milliseconds)
\begin{lstlisting}[language=SMLSmall]
SRW_TAC [] [SURJ_DEF, INJ_DEF] THEN METIS_TAC []
\end{lstlisting}

Goal
\begin{lstlisting}[language=SMLSmall]
!l1 l2 n x. LENGTH l1 <= n ==> 
  (LUPDATE x n (l1 ++ l2) = l1 ++ (LUPDATE x (n - LENGTH l1) l2))
\end{lstlisting}

@@

Original \holfour proof script (63 milliseconds)
\begin{lstlisting}[language=SMLSmall]
  rw[] THEN simp[LIST_EQ_REWRITE] THEN Q.X_GEN_TAC `z` THEN
  simp[EL_LUPDATE] THEN rw[] THEN simp[EL_APPEND2,EL_LUPDATE] THEN
  fs[] THEN Cases_on `z < LENGTH l1` THEN
  fs[] THEN simp[EL_APPEND1,EL_APPEND2,EL_LUPDATE]
\end{lstlisting}

@@

%val LUPDATE_APPEND2 = Q.store_thm("LUPDATE_APPEND2",
%   `!l1 l2 n x.
%      LENGTH l1 <= n ==>
%      (LUPDATE x n (l1 ++ l2) = l1 ++ (LUPDATE x (n-LENGTH l1) l2))`,
%);
\vspace{5mm}
Make it match the new version.
\tactictoe proof script (17 milliseconds)
\begin{lstlisting}[language=SMLSmall]
Induct_on `l1` THENL [SRW_TAC [] [],
  Cases_on `n` THENL [SRW_TAC [] [],
    FULL_SIMP_TAC (srw_ss ()) [] THEN METIS_TAC [LUPDATE_def]]]
\end{lstlisting}
fs is a \sml binding for FULL\_SIMP\_TAC (srw\_ss ()) and rw is short for ??.
